\section{Future Work}
\label{sec:future_work}

\paragraph{Rendez-vous}

The main limitation of transactions is that they are invisible to each other.
A transaction does not directly modify shared-memory locations and, once it does, the modifications appear as atomic to other transactions and outside observers.
The mutual invisibility means that \emph{rendez-vous} between two (or more) threads cannot be expressed as a pair of composable transactions.
For example, it is not possible to implement synchronous message passing as can be found, \eg, in \ConcurrentML, \Go, and various other languages and libraries.
However, it is possible~\citep{DBLP:conf/wotug/Brown08,DBLP:conf/icfp/DonnellyF06} to implement a compositional message passing model on top of STM.
Allowing synchronisation between threads requires committing multiple transactions.

\paragraph{Relaxed memory model}

The verification assumes a \emph{sequentially consistent} memory model.
We would like to extend it to the \emph{relaxed} memory model of \OCamlFive~\citep{DBLP:conf/pldi/DolanSM18}, which is not currently supported by the \Zoo framework.
Similarly to primitive atomic locations~\citep{DBLP:journals/pacmpl/MevelJP20}, MCAS locations would carry a logical \emph{memory view}.
$\c{get}\ \zooLoc$ would acquire the view stored in \zooLoc.
On success, $\c{cas}\ \zooLoc{}s\ \v{befores}\ \v{afters}$ would release the view of the caller through locations $\zooLoc{}s$ and acquire the views of the locations.

\paragraph{Verification of the full library}

We would like to extend the verification to the full \Kcas library, including its transactional API.
This is challenging for two reasons.
(1) The real implementation is even more complex: internally, CAS operations are stored in a splay tree rather than a list; the MCAS algorithm traverses the tree in a depth-first manner.
(2) Specifying transactions in a composable way appears to be non-trivial; we anticipate that it will require several layers of specifications.
That being said, the implementation is relatively short (approximatively one thousand lines of code), so it should be feasible in practice.
