\section{\Kcas Overview}
\label{sec:overview}

\Kcas provides two main abstractions: (1) \emph{shared-memory locations} can be created and manipulated using the \ocamlinline{Loc} module; (2) \emph{explicit transaction logs} allow writing transactions over shared-memory locations using the \ocamlinline{Xt} module.
The following sections discuss them in turn.

\subsection{Shared-Memory Locations}

\input{figures/loc_interface}

Similarly to \c{TVar} in \Haskell's STM, \Kcas provides an abstraction of shared-memory locations through the \ocamlinline{Loc} module (\cref{fig:loc_interface}).
It is essentially a superset of the \ocamlinline{Atomic} module from the standard library.
Thus, an application that uses \ocamlinline{Atomic} and needs to perform atomic operations over multiple locations can simply rebind \ocamlinline{module Atomic = Loc} and then use the \ocamlinline{Xt} API (see \cref{sec:overview_xt}).
This should not be done just-in-case, however, as it does naturally have higher overhead than \ocamlinline{Atomic}.

\subsection{Direct-Style Transactional Interface}
\label{sec:overview_xt}

\input{figures/xt_interface}

Transactions can be constructed and executed using the \ocamlinline{Xt} module (\cref{fig:xt_interface}).
In \Kcas, a transaction is essentially a function of type \ocamlinline{xt:Xt.t -> 'a}, where \ocamlinline{xt} represents the \emph{explicit transaction log}.
As a result, we get sequential and conjunctive composition of transactions for free:

\smallskip
\begin{minipage}[t]{.45\textwidth}
\begin{ocamlcode}
let bind ~xt tx1 tx2 =
  let x = tx1 ~xt in tx2 ~xt x
\end{ocamlcode}
\end{minipage}
\hfill
\begin{minipage}[t]{.45\textwidth}
\begin{ocamlcode}
let pair ~xt tx1 tx2 =
  (tx1 ~xt, tx2 ~xt)
\end{ocamlcode}
\end{minipage}
\smallskip

The log records transactional operations like \ocamlinline{Xt.get} and \ocamlinline{Xt.set}.
For example, we can write a transaction to remove a node from a doubly-linked circular list as follows:
\begin{ocamlcode}
let remove ~xt node =
  (* Read pointers to the predecessor and successor nodes: *)
  let pred = Xt.get ~xt node.pred and succ = Xt.get ~xt node.succ in
  (* Update pointers in this node: *)
  Xt.set ~xt node.succ node ; Xt.set ~xt node.pred node ;
  (* Update pointers to this node: *)
  Xt.set ~xt pred.succ succ ; Xt.set ~xt succ.pred pred
\end{ocamlcode}
To actually remove a node, we need to commit the transaction:
\begin{ocamlcode}
Xt.commit { tx = remove node }
\end{ocamlcode}

This \emph{direct} style is in contrast to the \emph{monadic} style of \Haskell's STM.
It has the advantage that it can easily be used with all the existing control flow structures of \OCaml that would need to be encoded with combinators in a monadic API, including conditionals, loops and higher-order functions like \ocamlinline{List.iter}.
On the other hand, a monadic API provides a very strict abstraction barrier against misuse by keeping users from accessing the log directly, which is not thread-safe and should not be accessed or reused after the transaction has been committed.

Fortunately, there is a way to make such misuse much more difficult by employing higher-rank polymorphism~\citep{DBLP:conf/pldi/LaunchburyJ94}: by adding a type variable to the type of the log and requiring a transaction to be universally quantified with respect to the log, the type system prevents a log from being reused.
It is still possible to \eg create a closure that refers to a log after it has been consumed, but that requires effort from the programmer and is unlikely to happen by accident.

In \Haskell's STM, the monadic API combined with the purity of the language gives a stronger guarantee~\citep{DBLP:conf/ppopp/HarrisMPH05}: a transaction can only perform memory operations through the transactional API.
In \OCaml, even with a monadic API, we cannot provide such a guarantee.
In practice, this means that it is the responsibility of the user not to do anything irrevocable during a transaction, \eg input/output effects.

\subsection{Scheduler-Friendly Blocking}
\label{sec:overview_blocking}

Similarly to \Haskell's STM, \Kcas provides a \emph{retry mechanism} that allows a transaction to wait for changes to any number of locations.
A transaction may simply raise the exception \ocamlinline{Retry.Later} to signal that it should be retried after another thread has made changes to the locations accessed so far.
The operations \ocamlinline{Loc.get_as}, \ocamlinline{Loc.update} and \ocamlinline{Loc.modify} also supports it.

A trivial example would be to convert a non-blocking queue operation to a blocking operation:
\begin{ocamlcode}
let take_blocking ~xt queue =
  match Queue.Xt.take_opt ~xt queue with
  | None -> Retry.later () (* equivalent to [raise Retry.Later]. *)
  | Some elem -> elem
\end{ocamlcode}

The main technical challenge in implementing the retry mechanism is that it should perform blocking in a scheduler-friendly manner.
In other words, other fibers, as in \Eio, or tasks, as in \Domainslib, should not be prevented from running while one of them is blocked.
The difficulty with that is that each scheduler potentially has its own way for suspending a fiber or waiting for a task.
To solve this problem, we use the \c{domain-local-await} library~\citep{domain-local-await}.
The idea is simple: schedulers like \Eio and \Domainslib install their own implementation of the blocking mechanism, stored in a domain-local variable, and then libraries like \Kcas can obtain the mechanism to block in a scheduler-friendly manner.
Concretely, this means \Kcas integrates well with existing and future schedulers.

\subsection{Scheduler-Friendly Timeouts}
\label{sec:overview_timeouts}

Blocking or retrying a transaction indefinitely is often not acceptable.
Fortunately, \Kcas provides direct support for \emph{timeouts}.
To perform a transaction with a timeout, one can simply explicitly specify a \ocamlinline{timeoutf} in seconds:
\begin{ocamlcode}
let try_take ~timeoutf queue =
  Xt.commit ~timeoutf { tx = Queue.Xt.take_blocking queue }
\end{ocamlcode}

Internally, \Kcas uses the \c{domain-local-timeout}~\citep{domain-local-timeout} library to implement timeouts.
Just like with the previously mentioned domain-local \emph{await}, the idea with domain-local \emph{timeout} is to allow libraries like \Kcas to tap into the native mechanism of whatever scheduler is currently in use and to do so conveniently without pervasive parameterization.

\subsection{Disjunctive and Conditional Composition}
\label{sec:overview_disjunctive_composition}

Another challenge is to support disjunctive composition of blocking transactions.
As explained by \citet{DBLP:conf/ppopp/HarrisMPH05}, this in turn requires support for nested transactions.

Consider the following attempt at a conditional blocking take from a queue:
\begin{ocamlcode}
let non_nestable_take_if ~xt predicate queue =
  let x = Queue.Xt.take_blocking ~xt queue in
  if not (predicate x) then Retry.later () ;
  x
\end{ocamlcode}
Now, suppose that we want to take an element from the first of two queues using this blocking transaction.
For that purpose, \Kcas provides the \ocamlinline{Xt.first} function: given a list of transactions \ocamlinline{txs}, \ocamlinline{Xt.first ~xt txs} calls each transaction in turn and either returns the value returned by the transaction or continues if the transaction raised \ocamlinline{Retry.Later}; if all the transactions raised \ocamlinline{Retry.Later}, so does \ocamlinline{Xt.first}.

However, \ocamlinline{Xt.first} does not automatically rollback changes made by the transactions.
Therefore, if one were to use the above transaction with \ocamlinline{Xt.first}, one would run into the following problem: while only a value that passes the predicate would be returned, an element might be taken from both queues.
To avoid this problem, we need a way to \emph{roll back changes} recorded by a transaction attempt.
The way \Kcas supports this is via an \emph{explicit scoping mechanism}.
Here is a working (nestable) version of conditional blocking take:
\begin{ocamlcode}
let take_if ~xt predicate queue =
  let snap = Xt.snapshot ~xt in
  let x = Queue.Xt.take_blocking ~xt queue in
  if not (predicate x) then Retry.later (Xt.rollback ~xt snap) ;
  x
\end{ocamlcode}
First a snapshot of the transaction log is taken and then, in case the predicate is not satisfied, a rollback to the snapshot is performed before signaling a retry.
The obvious disadvantage of this explicit approach is that it requires more care from the programmer.
The advantage is that it allows the programmer to explicitly scope nested transactions and perform rollbacks only when necessary and in a more fine-tuned manner, which can allow for better performance.

As a result, \Kcas also supports conditional composition:
\begin{ocamlcode}
let if_else ~xt predicate tx1 tx2 =
  let snap = Xt.snapshot ~xt in
  let x = tx1 ~xt in
  if predicate x then x else (Xt.rollback ~xt snap ; tx2 ~xt)
\end{ocamlcode}

\subsection{Read Skews}

\Kcas, and more precisely the underlying MCAS algorithm (see \cref{sec:implementation_mcas}), ensures that (1) it is \emph{not} possible to read uncommitted changes to shared-memory locations and (2) successful transactions are \emph{linearizable}, meaning a transaction can only commit after all of the accesses taken together have been atomic.

Unfortunately, \Kcas does not prevent one specific concurrency anomaly: when a transaction reads multiple locations, it is possible for the transaction to observe an inconsistent state when other transactions commit changes between reads of different locations.
This is traditionally called a \emph{read skew} in database terminology.
Several other STM implementations have the same property, \eg \Haskell's STM~\citep{DBLP:conf/ppopp/HarrisMPH05} and \ZIO~\citep{zio}.

Having observed such an inconsistent state, a transaction cannot succeed and must be retried.
However, even though the transaction must retry, read skews can still cause serious problems.
Consider, for example, the following transaction:
\begin{ocamlcode}
let unsafe_subscript ~xt arr idx =
  let a = Xt.get ~xt arr and i = Xt.get ~xt idx in
  (* Invariant between a and i may not hold. *)
  a.(i)
\end{ocamlcode}
The assumption is that the \ocamlinline{arr} and \ocamlinline{idx} locations are always updated atomically such that the subscript operation should be safe.
Unfortunately, due to read skews, the array and index might not match and the subscript operation could result in an ``index out of bounds'' exception.
Even more subtle problems are possible.
For example, a balanced binary search tree implementation using rotations can, due to read skews, be seen to have a cycle.

\Kcas addresses these problems in a couple of ways.
First, it performs periodic validation of the entire transaction log when an access, such as \ocamlinline{Xt.get} or \ocamlinline{Xt.set}, is made through the log.
As it would take quadratic time to validate the entire log on every access --- and therefore change the time complexity of transactions ---, the number of accesses between validations is doubled after each validation.
Periodic validation is an effective way to make loops that access shared memory locations, such as the lookup of a key from a binary search tree, resistant against read skews; such loops will eventually be aborted on some access and will then be retried.
Periodic validation, however, is \emph{not} effective against problems that might occur due to non-transactional operations made after reading inconsistent state and depending on invariants.
For those cases, an explicit \ocamlinline{Xt.validate} operation is provided that can be used to validate that the accesses to particular locations have been atomic:
\begin{ocamlcode}
let subscript ~xt arr idx =
  let a = Xt.get ~xt arr and i = Xt.get ~xt idx in
  (* Validate accesses after making them: *)
  Xt.validate ~xt arr ; Xt.validate ~xt idx ;
  (* Invariant between a and i now holds. *)
  a.(i)
\end{ocamlcode}

\subsection{Composable Concurrent Data Structures}

\Kcas comes with a companion library \KcasData that provides composable concurrent data structures: list, stack, queue, doubly-linked list, hash table.
The LRU cache from \cref{sec:introduction}, for example, can be implemented using the provided doubly-linked list and hash table.
We also utilize \KcasData for benchmarks (see \cref{sec:benchmarks}).
